A comparison of model-based and design-based approaches for spatial data.

### 2021-06-09 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
adjusted final intro paragraph
update to title

---
title: A comparison of model-based and design-based approaches for spatial data.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matthew Higham
    email: mhigham@stlaw.edu
    affiliation: STLAW
    footnote: 1
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Math, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

@@

The rest of this paper is organized as follows. In Section \ref{sec:background}, we compare sampling and estimation procedures between the design-based approach and the model-based approach. In Section \ref{sec:numstudy}, we use simulated and real data to compare the properties of parameter estimates from both aproaches. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

>>>>

---
title: A comparison of design-based and model-based approaches for spatial data.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matthew Higham
    email: mhigham@stlaw.edu
    affiliation: STLAW
    footnote: 1
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Math, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

@@

The rest of this paper is organized as follows. In Section \ref{sec:background}, we compare sampling and estimation procedures between the design-based approach and the model-based approach. In Section \ref{sec:numstudy}, we use simulated and real data to study the properties of parameter estimates from both approaches. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

### 2021-06-09 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
update to title

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as variables measured at specific
geographic locations. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered, but
they tend to compare design-based approaches that ignore spatial
location in sampling to model-based approaches (Brus and De Gruijter,
1997; Ver Hoef, 2002; Ver Hoef, 2008). More recent overviews include
Brus (2020) and (\textbf{wang2012spatial?}), but no numerical comparison
has been made between design-based approaches that incorporate spatial
location in sampling and model-based approaches.

The rest of this paper is organized as follows. In Section
\ref{sec:background}, we compare sampling and estimation procedures
between the design-based approach and the model-based approach. In
Section \ref{sec:numstudy}, we use simulated and real data to compare
the properties of parameter estimates from both aproaches. And in
Section \ref{sec:discussion}, we end with a discussion and provide
directions for future research.

@@

Several comparisons between the two paradigms have been made in many
different contexts. Sterba (2009) give some history of the two paradigms
as well as some applications in psychological contexts. Cooper (2006)
review the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g. Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-barabesi2011sampling}{}%
Barabesi, L., Franceschi, S., 2011. Sampling properties of spatial total
estimators under tessellation stratified designs. Environmetrics 22,
271--278.

@@

\leavevmode\hypertarget{ref-chiles1999geostatistics}{}%
Chiles, J.-P., Delfiner, P., 1999. Geostatistics: {Modeling Spatial
Uncertainty}. {John Wiley \& Sons}, New York.

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-NN) methods for forestry applications. PloS
one 8, e59129.

@@

\end{CSLReferences}

>>>>

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as variables measured at specific
geographic locations. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered, but
they tend to compare design-based approaches that ignore spatial
location in sampling to model-based approaches (Brus and De Gruijter,
1997; Ver Hoef, 2002, 2008). More recent overviews include Brus (2020)
and ({\textbf{???}}), but no numerical comparison has been made between
design-based approaches that incorporate spatial location in sampling
and model-based approaches.

The rest of this paper is organized as follows. In Section
\ref{sec:background}, we compare sampling and estimation procedures
between the design-based approach and the model-based approach. In
Section \ref{sec:numstudy}, we use simulated and real data to study the
properties of parameter estimates from both approaches. And in Section
\ref{sec:discussion}, we end with a discussion and provide directions
for future research.

@@

Several comparisons between the two paradigms have been made in many
different contexts. Sterba (2009) give some history of the two paradigms
as well as some applications in psychological contexts. Cooper (2006)
review the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g.~Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-barabesi2011sampling}{}%
Barabesi, L., Franceschi, S., 2011. Sampling properties of spatial total
estimators under tessellation stratified designs. Environmetrics 22,
271--278.

@@

\leavevmode\hypertarget{ref-chiles1999geostatistics}{}%
Chiles, J.-P., Delfiner, P., 1999. Geostatistics: Modeling Spatial
Uncertainty. John Wiley \& Sons, New York.

@@

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-nn) methods for forestry applications. PloS
one 8, e59129.

@@
